---
title: "MLE,EM,EM acceleration"
author: "Guanting Wei"
date: "Nov.3.2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Suppose that there are 4 kinds of animals(A,B,C,D) in a farm. And their probability are $$P(A)=\frac{1}{2}+\frac{\theta}{4}$$ 
$$P(B)=\frac{1}{4}(1-\theta)$$ 
$$P(C)=\frac{1}{4}(1-\theta)$$ 
$$P(D)=\frac{\theta}{4}$$
$$\theta \in(0,1)$$. 
Now we randomly choose one animal 197 times and get the result
$$N(A)=125$$
$$N(B)=18$$
$$N(C)=20$$
$$N(D)=34$$

# MLE
We let $X=(x_1,x_2,x_3,x_4)=(125,18,20,34)$
\[
\begin{aligned}
L(x_i;\theta)&=(\frac{1}{2}+\frac{\theta}{4})^{x_1}[\frac{1}{4}(1-\theta)]^{x_2}[\frac{1}{4}(1-\theta)]^{x_3}(\frac{\theta}{4})^{x_4}\\
&\propto (2+\theta)^{x_1}(1-\theta)^{x_2+x_3}(\theta)^{x_4}
\end{aligned}
\]
\[
\begin{aligned}
l(x_i;\theta)&=\ln[(2+\theta)^{x_1}(1-\theta)^{x_2+x_3}(\theta)^{x_4}]\\&=x_1\ln(2+\theta)+(x_2+x_3)\ln(1-\theta)+x_4\ln{\theta}\\&=125\ln(2+\theta)+38\ln(1-\theta)+34\ln{\theta}
\end{aligned}
\]
\[
\begin{aligned}
l'(x_i;\theta)=\frac{125}{2+\theta}-\frac{38}{1-\theta}+\frac{34}{\theta}=0
\end{aligned}
\]
We use Newton’s Method to find the result
```{r}
f=function(x){
  125/(2+x)-38/(1-x)+34/x
}
f1=function(x){
  -125/(2+x)^2-38/(1-x)^2-34/x^2
}
x=0.1
count=0
  while (abs(f(x))>0.001&&count<1000) {
    temp=x
    x=temp-f(temp)/f1(temp)
    count=count+1
  }
x

```
We get $\theta=0.6268216$

# EM
Let's suppose that we have two subspecies of A including $A_1$,$A_2$ and 
$$P(A_1)=\frac{1}{2}$$
$$P(A_2)=\frac{\theta}{4}$$
Further more, suppose that
$$N(A_1)=Y$$
$$N(A_2)=x_1-Y=125-Y$$
In this situation, if we want to use MLE to get $\theta$,we will have
\[
\begin{aligned}
L(X,Y;\theta)&=(\frac{1}{2})^Y(\frac{\theta}{4})^{x_1-Y}[\frac{1}{4}(1-\theta)]^{x_2}[\frac{1}{4}(1-\theta)]^{x_3}(\frac{\theta}{4})^{x_4}\\
&\propto (1-\theta)^{x_2+x_3}(\theta)^{x_1-Y+x_4}
\end{aligned}
\]
We can not calculate $\theta$ through this formula.
Then we need EM algorithm
\[
\begin{aligned}
Q(\theta|{\theta}^{(i)})&=E[\ln L(X,Y;\theta|X,{\theta}^{(i)})]\\&
=E[(x_1-Y+x_4)\ln \theta+(x_2+x_3)\ln (1-\theta)|X,{\theta}^{(i)}]\\&
=(x_1+x_4)\ln \theta+(x_2+x_3)\ln (1-\theta)-E(Y|X,{\theta}^{(i)})\ln \theta
\end{aligned}
\]
When given $X$ and ${\theta}^{(i)}$, $(Y|X,{\theta}^{(i)})\sim b(x_1,p_1)$,where $p_1=\frac{\frac{1}{2}}{\frac{1}{2}+\frac{{\theta}^{(i)}}{4}}=\frac{2}{2+{\theta}^{(i)}}$
$$E(Y|X,{\theta}^{(i)})=np=x_1p_1=\frac{2x_1}{2+{\theta}^{(i)}}$$
So
\[
\begin{aligned}
Q(\theta|{\theta}^{(i)})&=(x_1+x_4)\ln \theta+(x_2+x_3)\ln (1-\theta)-E(Y|X,{\theta}^{(i)})\ln \theta\\&=(x_1+x_4-\frac{2x_1}{2+{\theta}^{(i)}})\ln \theta+(x_2+x_3)\ln (1-\theta)\\&=
(159-\frac{250}{2+{\theta}^{(i)}})\ln \theta+38\ln (1-\theta)
\end{aligned}
\]
Then we calculate the maximum of $Q(\theta|{\theta}^{(i)})$
\[
\begin{aligned}
&Q'(\theta|{\theta}^{(i)})=(159-\frac{250}{2+{\theta}^{(i)}})\frac{1}{\theta}-38\frac{1}{(1-\theta)}=0\\&
\Rightarrow {\theta}^{(i+1)}=\frac{68+159{\theta}^{(i)}}{144+197{\theta}^{(i)}}
\end{aligned}
\]

```{r}
EM=function(theta.init,tol,count.max){
  theta=theta.init
  count=0
  for(i in 1:count.max){
    count=count+1
    theta[i+1]=(68+159*theta[i])/(144+197*theta[i])
  if(abs(theta[i+1]-theta[i])<tol&&count<count.max)break
  }
  theta
}
theta=EM(0.2,1e-5,1000)
count=length(theta)-1
theta
count
```

# MCEM
We can also use Monte Carlo Method to do this EM algorithm.
\[
\begin{aligned}
Q^{(i+1)}(\theta|\theta^{(i)})&=\frac{1}{m}\sum_{j=1}^{m}\ln{L(X,Y;\theta)}\\&
=\frac{1}{m}\sum_{j=1}^{m}\ln(\theta^{x_1+x_2-Y}(1-\theta)^{x_2+x_3})\\&
=\frac{1}{m}\sum_{j=1}^{m}(x_1+x_2-Y)\ln\theta+(x_2+x_3)\ln(1-\theta)\\&
=(x_1+x_2-E(Y))\ln\theta+(x_2+x_3)\ln(1-\theta)
\end{aligned}
\]

\[
\begin{aligned}
&Q'^{(i+1)}(\theta|\theta^{(i)})=\frac{(x_1+x_2-E(Y))}{\theta}-\frac{(x_2+x_3)}{1-\theta}=0\\&
\Rightarrow \theta^{(i+1)}=\frac{x_1+x_4-E(Y)}{x_1+x_2+x_3+x_4-E(Y)}\\&
Y\sim b(x_1,\frac{2}{2+{\theta}^{(i)}})
\end{aligned}
\]


```{r}
x1=125
x2=18
x3=20
x4=34
MCEM=function(theta.init,m,tol,count.max){
  theta=theta.init
  count=0
  for(i in 1:count.max){
    count=count+1
    p1=2/(2+theta.init)
    Y=rbinom(m,x1,p1)
    theta[i+1]=(x1+x4-mean(Y))/(x1+x2+x3+x4-mean(Y))
    theta.init=theta[i+1]
    if(abs(theta[i+1]-theta[i])<tol&&count<count.max)break
  }
  theta
}
theta=MCEM(0.2,10000,1e-5,1000)
count=length(theta)-1
theta
count
```


# SQUAREM
```{r}
x1=125
x2=18
x3=20
x4=34
loglik=function(theta){
  y=rbinom(1,x1,2/(2+theta))
  log((1-theta)^(x2+x3)*theta^(x1+x4-y))
  }
em=function(theta){
  theta.new=(68+159*theta)/(144+197*theta)
  theta=theta.new
  return(theta)
}
library(SQUAREM)
theta0=0.2
result=squarem(c(theta=theta0), fixptfn=em, objfn=loglik, control=list(tol=1e-10,mstep=5))
theta=result$par
count=result$iter
theta
count
```

# Compare
Let's choose a smaller tol to see the speed of the algorithms we talk above.
```{r}
tol=1e-7
count.max=1000
theta.init=0.2
```

## EM
```{r}
t1=Sys.time()
theta_EM=EM(theta.init,tol,count.max)
t2=Sys.time()
iter_EM=length(theta_EM)-1
theta_EM
print(paste("iternation=",iter_EM))
print(paste("time=",t2-t1))
```

## MCEM
```{r}
t3=Sys.time()
theta_MCEM=MCEM(theta.init,10000,tol,count.max)
t4=Sys.time()
iter_MCEM=length(theta_MCEM)-1
print("MCEM")
theta_MCEM
print(paste("iternation=",iter_MCEM))
print(paste("time=",t4-t3))
```

## SQUAREM
```{r}
t5=Sys.time()
result=squarem(c(theta=theta0), fixptfn=em, objfn=loglik, control=list(tol=1e-10,mstep=5))
t6=Sys.time()
theta_SQUAREM=result$par
iter_SQUAREM=result$iter
print("SQUAREM")
theta_SQUAREM
print(paste("iternation=",iter_SQUAREM))
print(paste("time=",t6-t5))
```

